Follow-up Samples

### 2017-01-13 submission/manuscript_outline_20161024.Rmd
add important variable component to results

***Bacterial Community and Fit Changes before and after Treatment***
Based on thetayc distance metrics, comparing the initial to the follow up samples, there was a significant difference between the adenoma and carcinoma groups (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a]. There was also a significant difference in FIT between initial and follow up samples with the carcinoma group having a significant decrease in FIT versus the adenoma group (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].  The whole community structure before and after surgery are visualized on NMDS graphs for both adenoma [Figure1c] (PERMANOVA = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) and carcinoma [Figure 1d] (PERMANOVA = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  When all initial and follow up samples were compared to each other there was no significant overall difference between them (PERMANOVA = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  There was also no significant difference between initial and follow up samples for observed OTUs, Shannon diversity, or evenness after correction for multiple comparisons [Table S1].  Time of follow up sample from initial sampling, did not have a significant difference between adenoma and carcinoma (uncorrected P-value = `r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)`).

@@

The test set AUC range for the full and reduced lesion model were similar to that reported for the training set AUC ranges and the ROC curve ranges overlap each other [Figure 3a].  The ROC curve for the final lesion model used falls within the range of both the full and reduced lesion model [Figure 3a].

***Most Important Variables to the Model***
Overall, there were a total of `r format(length(rownames(imp_model_vars)))` variables identified as being present in more than 50% of the training models [Figure S2]. The top 5 most important bacterial OTUs were `r paste(rf_lowest_ID[1], " (", names(rf_lowest_ID)[1], ")", sep = "")`,  `r paste(rf_lowest_ID[2], " (", names(rf_lowest_ID)[2], ")", sep = "")`, `r paste(rf_lowest_ID[3], " (", names(rf_lowest_ID)[3], ")", sep = "")`, `r paste(rf_lowest_ID[4], " (", names(rf_lowest_ID)[4], ")", sep = "")`, and `r paste(rf_lowest_ID[5], " (", names(rf_lowest_ID)[5], ")", sep = "")`.  These 5 OTUs were present in at least `r min(filter(imp_model_vars, Variable %in% names(rf_lowest_ID)[1:5])[, "total_appearance"])` out of the total `r length(rownames(model_train_summary))` different 80/20 runs.

***Surgical Removal of an Adenoma or Carcinoma Results in a Decrease in Positive Probability Prediction***
A total of `r length(rownames(good_metaf)) - length(rownames(filter(good_metaf, !is.na(fit_followUp))))` sample was omitted from the original `r length(rownames(good_metaf))` test sample set since it was missing a complete set of follow up data.  This left a total of `r sum(initial_cm$ref_yes)` samples for test predictions.  After multiple comparison correction there was a significant overall decrease in positive probability of a carcinoma and adenoma (P-value = `r format(IF_model_wilcox_paired_summary["lesion", "BH_correction"], scientific = TRUE, digits = 3)`) [Figure 4].  This decrease was significant for both adenoma (P-value = `r format(IF_model_wilcox_paired_summary["all_adenoma", "BH_correction"], scientific = TRUE, digits = 3)` [Figure 4a] and carcinoma (P-value = `r format(IF_model_wilcox_paired_summary["carcinoma_only", "BH_correction"], scientific = TRUE, digits = 3)`) [Figure 4b] alone.  This decrease in probability of lesion also held specifically for those with screen relative neoplasias (SRN) (P-value = `r format(IF_model_wilcox_paired_summary["SRN_only", "BH_correction"], scientific = TRUE, digits = 3)`).  A total of `r initial_cm["pred_yes", "ref_yes"]` or `r format(confusion_summary["Accuracy", "initial"]*100, digits = 3)`% of all samples were correctly predicted to have a lesion.  Although there was a decrease in positive probability only `r followup_cm["pred_no", "ref_no"]` of the total `r sum(initial_cm$ref_yes)` individuals were classified as adenoma or carcinoma free on follow up (successful classification of `r format(confusion_summary["Accuracy", "followup"]*100, digits = 3)`%). There was no significant difference between the predictions and actual diagnosis for the initial samples in the `r length(rownames(good_metaf))`-sample cohort test set.  However, the predictions were significantly discordent with the diagnosis for the follow up samples (P-value = `r format(confusion_summary["McnemarPValue", "followup"], scientific = TRUE, digits = 3)`).  Although there were discordent results the respective sensitivity for the initial group was `r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)`% and for follow up was `r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)`%, respectively.

There was `r length(rownames(filter(good_metaf, Disease_Free == "n")))` individual who still clearly had CRC on follow up as well as `r length(rownames(filter(good_metaf, Disease_Free == "unknown")))` individuals whose status on follow up was unknown.  Although the `r length(rownames(filter(good_metaf, Disease_Free == "n")))` individual had a decrease in positive probability their follow up sample was still higher than the cutoff threshold of 0.5 (positive probability = `r format(filter(follow_up_prob_summary, sampleType == "followup" & disease_free == "n")[, "Yes"], digits = 3)`).  Interestingly, `r length(rownames(filter(follow_up_prob_summary, sampleType == "followup" & disease_free == "unknown" & Yes > 0.5)))` individuals who were unknown on follow up still were over the threshold cutoff of 0.5 even though, like the `r length(rownames(filter(good_metaf, Disease_Free == "n")))` individual with clear CRC on follow up, the probability of an adenoma or carcinoma decreased [Table S2].

The follow up positive probabilities were not affected by either chemotherapy treatment (uncorrected P-value = `r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "followup", chemo == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "followup", chemo == "no")[, "Yes"])$p.value, digits = 3)`) or  radiation therapy (uncorrected P-value = `r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "followup", rads == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "followup", rads == "no")[, "Yes"])$p.value, digits = 3)`).  There was also no difference in the amount of change in the positive probability based on whether individuals received chemotherapy (unccorected P-value = `r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "initial", chemo == "yes")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", chemo == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "initial", chemo == "no")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", chemo == "no")[, "Yes"])$p.value, digits =3)`) or radiation therapy (uncorrected P-value = `r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "initial", rads == "yes")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", rads == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "initial", rads == "no")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", rads == "no")[, "Yes"])$p.value, digits = 3)`).

>>>>

***Bacterial Community and Fit Changes before and after Treatment***
Based on thetayc distance metrics, comparing the initial to the follow up samples, there was a significant difference between the adenoma and carcinoma groups (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a]. There was also a significant difference in FIT between initial and follow up samples with the carcinoma group having a significant decrease in FIT versus the adenoma group (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].  The whole community structure before and after surgery are visualized on NMDS graphs for both adenoma [Figure1c] (PERMANOVA = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) and carcinoma [Figure 1d] (PERMANOVA = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  When all initial and follow up samples were compared to each other there was no significant overall difference between them (PERMANOVA = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  There was also no significant difference between initial and follow up samples for observed OTUs, Shannon diversity, or evenness after correction for multiple comparisons [Table S1].  There was no significant difference between initial and follow up samples for any single OTU [Figure S1].

@@

The test set AUC range for the full and reduced lesion model were similar to that reported for the training set AUC ranges and the ROC curve ranges overlap each other [Figure 3a].  The ROC curve for the final lesion model used falls within the range of both the full and reduced lesion model [Figure 3a].  Interestingly, the test set AUC range for the initial and follow up performed much better then the training set AUCs.  Both the full and reduced initial and follow up models overlapped with each other [Figure 3b] there was a marked decrease in the ROC curve for the final model used.

***Most Important Variables to the Models***
The reduced models were built based on the most important variables to the respective full model.  For the lesion model there were a total of `r length(lesion_top10_varsMDA$variable)` variables [Figure S2] whereas for the initial and follow up model there were a total of `r length(IF_top10_varsMDA$variable)` variables [Figure S3].  For both models FIT resulted in the largest decrease in MDA [Figure S2a & S3a].

***Positive Probability Prediction after Surgical Removal of Adenoma or Carcinoma***

**Treatment and Time Differences**

There was also no difference in the amount of change in the positive probability based on whether individuals received chemotherapy (unccorected P-value = `r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "initial", chemo == "yes")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", chemo == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "initial", chemo == "no")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", chemo == "no")[, "Yes"])$p.value, digits =3)`) or radiation therapy (uncorrected P-value = `r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "initial", rads == "yes")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", rads == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "initial", rads == "no")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", rads == "no")[, "Yes"])$p.value, digits = 3)`). Time of follow up sample from initial sampling, did not have a significant difference between adenoma and carcinoma (uncorrected P-value = `r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)`).
